Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-06-16 writing/DraftExposurePaper/DraftExposurePaper.Rmd
Add to writing draft.

Many outcomes of interest are available at county-level aggregations (e.g., counts of health outcomes: @Grabich2016: premature births and low birth weights). Further, decisions and policies to prepare for and respond to storms are often undertaken at the county level [@Zandbergen2009]. Some studies suggest methods to determine exposure, including inland exposure, to tropical storms at the county-level over many years [@Zandbergen2009]. Here, we offer tools, through the open-source `hurricaneexposure` package, to determine county-level exposure to tropical storms using several different metrics (distance, rainfall, wind). In this paper, we describe how this software package assesses exposure to tropical storms at the county level, provide an analysis of the advantages and disadvantages of different strategies for assessing county-level exposure, and explore variation in exposure assessments based on using different exposure metrics.

Current hurricane exposure assignments are often nonspecific [GBA-- I found this wording a bit unclear] and may lead to missclassification of exposure at the county-level [@Grabich2016]. Researchers have found heterogeneity in exposure assignments of counties, both within and between storms, when different metrics of exposure are used [@Grabich2016]. Such missclassification can have important implications of assessments of the public health and economic impacts of tropical storms.

Many of the studies that have assessed exposure to tropical storms in the US have used geographical information system software (e.g., ArcGIS) [example studies]. Here, we offer methods to map and output historic exposure to tropical storms that does not require the use of proprietary software but instead uses a package written in the R statistical programming language, which is free and open-source.

@@

A few studies have sought to determine exposure to tropical storms over a  study period that includes multiple storms. For example, ... assessed the exposure of counties in Florida to hurricanes during the 2004 season [citation]. The largest county-level exposure study is likely that of @Zandbergen2009, which estimated exposure in US counties to all [US landfalling? or all storms?] Atlantic basin tropical storms between 1851 and 2003, using both distance and an exposure metric that incorporated distance and windspeed. They used this exposure evaluation to create maps of total exposure to tropical storms within US counties, as well as to explore associations between a county's long-term exposure to tropical storms and its location, distance from the coast, size, and shape [@Zandbergen2009].

@@

Another study used distance buffers of 30, 60, and 100 kilometers for different metrics of distance-based exposure [@Grabich2016] [check the paper-- was this distance to any part of a county, or to the center of the county?].

@@

@@

Studies that have used wind-based exposure metrics have used a variety of windspeed cutoff values to determine exposure. One study, for example, assigned a binary exposure value based on windspeed thresholds of 63 and 119 kilometers per hour, as well as a continuous exposure metric based on the exact value of the windspeed in the county during the storm and a categorical exposure metric based on the Saffir-Simpson storm severity categories [@Grabich2016].

@@

>>>>

Many outcomes of interest are available at county-level aggregations (e.g., counts of health outcomes: @Grabich2016: premature births and low birth weights; @Grabich2015: birth rates). Further, decisions and policies to prepare for and respond to storms are often undertaken at the county level [@Zandbergen2009]. Some studies suggest methods to determine exposure, including inland exposure, to tropical storms at the county-level over many years [@Zandbergen2009]. Here, we offer tools, through the open-source `hurricaneexposure` package, to determine county-level exposure to tropical storms using several different metrics (distance, rainfall, wind). In this paper, we describe how this software package assesses exposure to tropical storms at the county level, provide an analysis of the advantages and disadvantages of different strategies for assessing county-level exposure, and explore variation in exposure assessments based on using different exposure metrics.

Current hurricane exposure assignments are often nonspecific [GBA-- I found this wording a bit unclear] and may lead to missclassification of exposure at the county-level [@Grabich2016]. Researchers have found heterogeneity in exposure assignments of counties, both within and between storms, when different metrics of exposure are used [@Grabich2016; @Grabich2015]. Such missclassification can have important implications of assessments of the public health and economic impacts of tropical storms.

Many of the studies that have assessed exposure to tropical storms in the US have used geographical information system software (e.g., ArcGIS) [example studies: @Grabich2016; @Zandbergen2009]. Here, we offer methods to map and output historic exposure to tropical storms that does not require the use of proprietary software but instead uses a package written in the R statistical programming language, which is free and open-source.

@@

A few studies have sought to determine exposure to tropical storms over a  study period that includes multiple storms. For example, @Grabich2016 assessed the exposure of counties in Florida to land-falling hurricanes during the 2004 season. The largest-scale county-level exposure study is likely that of @Zandbergen2009, which estimated exposure in US counties to all [US landfalling? or all storms?] Atlantic basin tropical storms between 1851 and 2003, using both distance and an exposure metric that incorporated distance and windspeed. They used this exposure evaluation to create maps of total exposure to tropical storms within US counties, as well as to explore associations between a county's long-term exposure to tropical storms and its location, distance from the coast, size, and shape [@Zandbergen2009].

Other studies have used county-level exposure criteria to determine exposure to specific storms. For example, @Grabich2015 determined county-level exposure to two severe storms in 2004 in Florida counties for an epidemiological study.

@@

Another study used distance buffers of 30, 60, and 100 kilometers for different metrics of distance-based exposure [@Grabich2016] [check the paper-- was this distance to any part of a county, or to the center of the county?]. Another study used a distance criterion of 60 kilometers from the storm track [@Grabich2015] [again, check to see if this was to any point in the county or to some center point of the county].

@@

Most distance exposure definitions in the past have been based on maximum wind radii based on maximum sustained winds [@Kruk2010].

@@

Studies that have used wind-based exposure metrics have used a variety of windspeed cutoff values to determine exposure. One study, for example, assigned a binary exposure value based on windspeed thresholds of 63 and 119 kilometers per hour, as well as a continuous exposure metric based on the exact value of the windspeed in the county during the storm and a categorical exposure metric based on the Saffir-Simpson storm severity categories [@Grabich2016]. Another study used a criterion of wind speed at or above 74 miles per hour [@Grabich2015].

@@

```{r echo = FALSE}
studies <- c("Czajkowski 2010", "Rezempour 2014", "Zandbergen 2009", 
             "Currie 2013", "Grabich 2016", "Zaharan 2010", "Kruk 2010", 
             "Villarini 2011", "Atallah 2007")
distance_km <- c("120", "", "Buffer", "25, 30, 60, 75", "30, 60, 100", 
                 "2 counties", "Radial Max Wind", "100-500", "")
rain_mm <- c("", "2/hr, 48/d", "", "", "", "", "", "", "200-500 (Floyd)")
wind_kt <- c("", "", "SSWS", "", "SSWS", "", "34, 50, 64", "", "")
exposure_parameters <- data.frame(studies, distance_km, rain_mm, wind_kt)
kable(exposure_parameters, "markdown")
```

SSWS = Staffir-Simpson Hurricane Wind Scale (TS, 1, 2, 3, 4, and 5). Wind Cutoffs at 34, 64, 83, 96, 113, and 137 kt.

### 2016-06-16 writing/DraftExposurePaper/DraftExposurePaper.tex
Add to writing draft.

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Assessing exposure to Atlantic Basin tropical storms in United States counties},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

Many outcomes of interest are available at county-level aggregations
(e.g., counts of health outcomes: Grabich et al. (2016): premature
births and low birth weights). Further, decisions and policies to
prepare for and respond to storms are often undertaken at the county
level (Zandbergen 2009). Some studies suggest methods to determine
exposure, including inland exposure, to tropical storms at the
county-level over many years (Zandbergen 2009). Here, we offer tools,
through the open-source \texttt{hurricaneexposure} package, to determine
county-level exposure to tropical storms using several different metrics
(distance, rainfall, wind). In this paper, we describe how this software
package assesses exposure to tropical storms at the county level,
provide an analysis of the advantages and disadvantages of different
strategies for assessing county-level exposure, and explore variation in
exposure assessments based on using different exposure metrics.

Current hurricane exposure assignments are often nonspecific {[}GBA-- I
found this wording a bit unclear{]} and may lead to missclassification
of exposure at the county-level (Grabich et al. 2016). Researchers have
found heterogeneity in exposure assignments of counties, both within and
between storms, when different metrics of exposure are used (Grabich et
al. 2016). Such missclassification can have important implications of
assessments of the public health and economic impacts of tropical
storms.

Many of the studies that have assessed exposure to tropical storms in
the US have used geographical information system software (e.g., ArcGIS)
{[}example studies{]}. Here, we offer methods to map and output historic
exposure to tropical storms that does not require the use of proprietary
software but instead uses a package written in the R statistical
programming language, which is free and open-source.

@@

A few studies have sought to determine exposure to tropical storms over
a study period that includes multiple storms. For example, \ldots{}
assessed the exposure of counties in Florida to hurricanes during the
2004 season {[}citation{]}. The largest county-level exposure study is
likely that of Zandbergen (2009), which estimated exposure in US
counties to all {[}US landfalling? or all storms?{]} Atlantic basin
tropical storms between 1851 and 2003, using both distance and an
exposure metric that incorporated distance and windspeed. They used this
exposure evaluation to create maps of total exposure to tropical storms
within US counties, as well as to explore associations between a
county's long-term exposure to tropical storms and its location,
distance from the coast, size, and shape (Zandbergen 2009).

@@

Another study used distance buffers of 30, 60, and 100 kilometers for
different metrics of distance-based exposure (Grabich et al. 2016)
{[}check the paper-- was this distance to any part of a county, or to
the center of the county?{]}.

@@

@@

Studies that have used wind-based exposure metrics have used a variety
of windspeed cutoff values to determine exposure. One study, for
example, assigned a binary exposure value based on windspeed thresholds
of 63 and 119 kilometers per hour, as well as a continuous exposure
metric based on the exact value of the windspeed in the county during
the storm and a categorical exposure metric based on the Saffir-Simpson
storm severity categories (Grabich et al. 2016).

@@

@@

\hypertarget{ref-Grabich2016}{}
Grabich, S. C., J. Horney, C. Konrad, and D. T. Lobdell. 2016.
``Measuring the Storm: Methods of Quantifying Hurricane Exposure with
Pregnancy Outcomes.'' \emph{Natural Hazards Review} 17 (1):
06015002--1--06015002--7.
doi:\href{https://doi.org/10.1061/(ASCE)NH.1527-6996.0000204}{10.1061/(ASCE)NH.1527-6996.0000204}.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Assessing exposure to Atlantic Basin tropical storms in United States counties},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

Many outcomes of interest are available at county-level aggregations
(e.g., counts of health outcomes: Grabich et al. (2016): premature
births and low birth weights; Grabich et al. (2015): birth rates).
Further, decisions and policies to prepare for and respond to storms are
often undertaken at the county level (Zandbergen 2009). Some studies
suggest methods to determine exposure, including inland exposure, to
tropical storms at the county-level over many years (Zandbergen 2009).
Here, we offer tools, through the open-source \texttt{hurricaneexposure}
package, to determine county-level exposure to tropical storms using
several different metrics (distance, rainfall, wind). In this paper, we
describe how this software package assesses exposure to tropical storms
at the county level, provide an analysis of the advantages and
disadvantages of different strategies for assessing county-level
exposure, and explore variation in exposure assessments based on using
different exposure metrics.

Current hurricane exposure assignments are often nonspecific {[}GBA-- I
found this wording a bit unclear{]} and may lead to missclassification
of exposure at the county-level (Grabich et al. 2016). Researchers have
found heterogeneity in exposure assignments of counties, both within and
between storms, when different metrics of exposure are used (Grabich et
al. 2016; Grabich et al. 2015). Such missclassification can have
important implications of assessments of the public health and economic
impacts of tropical storms.

Many of the studies that have assessed exposure to tropical storms in
the US have used geographical information system software (e.g., ArcGIS)
(example studies: Grabich et al. 2016; Zandbergen 2009). Here, we offer
methods to map and output historic exposure to tropical storms that does
not require the use of proprietary software but instead uses a package
written in the R statistical programming language, which is free and
open-source.

@@

A few studies have sought to determine exposure to tropical storms over
a study period that includes multiple storms. For example, Grabich et
al. (2016) assessed the exposure of counties in Florida to land-falling
hurricanes during the 2004 season. The largest-scale county-level
exposure study is likely that of Zandbergen (2009), which estimated
exposure in US counties to all {[}US landfalling? or all storms?{]}
Atlantic basin tropical storms between 1851 and 2003, using both
distance and an exposure metric that incorporated distance and
windspeed. They used this exposure evaluation to create maps of total
exposure to tropical storms within US counties, as well as to explore
associations between a county's long-term exposure to tropical storms
and its location, distance from the coast, size, and shape (Zandbergen
2009).

Other studies have used county-level exposure criteria to determine
exposure to specific storms. For example, Grabich et al. (2015)
determined county-level exposure to two severe storms in 2004 in Florida
counties for an epidemiological study.

@@

Another study used distance buffers of 30, 60, and 100 kilometers for
different metrics of distance-based exposure (Grabich et al. 2016)
{[}check the paper-- was this distance to any part of a county, or to
the center of the county?{]}. Another study used a distance criterion of
60 kilometers from the storm track (Grabich et al. 2015) {[}again, check
to see if this was to any point in the county or to some center point of
the county{]}.

@@

Most distance exposure definitions in the past have been based on
maximum wind radii based on maximum sustained winds (Kruk et al. 2010).

@@

Studies that have used wind-based exposure metrics have used a variety
of windspeed cutoff values to determine exposure. One study, for
example, assigned a binary exposure value based on windspeed thresholds
of 63 and 119 kilometers per hour, as well as a continuous exposure
metric based on the exact value of the windspeed in the county during
the storm and a categorical exposure metric based on the Saffir-Simpson
storm severity categories (Grabich et al. 2016). Another study used a
criterion of wind speed at or above 74 miles per hour (Grabich et al.
2015).

@@

\begin{longtable}[]{@{}llll@{}}
\toprule
studies & distance\_km & rain\_mm & wind\_kt\tabularnewline
\midrule
\endhead
Czajkowski 2010 & 120 & &\tabularnewline
Rezempour 2014 & & 2/hr, 48/d &\tabularnewline
Zandbergen 2009 & Buffer & & SSWS\tabularnewline
Currie 2013 & 25, 30, 60, 75 & &\tabularnewline
Grabich 2016 & 30, 60, 100 & & SSWS\tabularnewline
Zaharan 2010 & 2 counties & &\tabularnewline
Kruk 2010 & Radial Max Wind & & 34, 50, 64\tabularnewline
Villarini 2011 & 100-500 & &\tabularnewline
Atallah 2007 & & 200-500 (Floyd) &\tabularnewline
\bottomrule
\end{longtable}

SSWS = Staffir-Simpson Hurricane Wind Scale (TS, 1, 2, 3, 4, and 5).
Wind Cutoffs at 34, 64, 83, 96, 113, and 137 kt.

@@

\hypertarget{ref-Grabich2016}{}
Grabich, S. C., J. Horney, C. Konrad, and D. T. Lobdell. 2016.
``Measuring the Storm: Methods of Quantifying Hurricane Exposure with
Pregnancy Outcomes.'' \emph{Natural Hazards Review} 17 (1): 06015002.
doi:\href{https://doi.org/10.1061/(ASCE)NH.1527-6996.0000204}{10.1061/(ASCE)NH.1527-6996.0000204}.

\hypertarget{ref-Grabich2015}{}
Grabich, S. C., Whitney R. Robinson, Stephanie M. Engel, Charles E.
Konrad, C. Konrad, and J. Horney. 2015. ``County-Level Hurricane
Exposure and Birth Rates: Application of Difference-in-Differences
Analysis for Confounding Control.'' \emph{Emerging Themes in
Epidemiology} 12: 19.
doi:\href{https://doi.org/10.1186/s12982-015-0042-7}{10.1186/s12982-015-0042-7}.
